Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-22 report/chapter03.Rnw
.

This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme. Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable. Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space. That alignment keeps objectives, diagnostics, and reported log-densities interoperable.

\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}

Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
\begin{equation}
  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
  \qquad \sigma_k > 0,
  \label{eq:standardization}
\end{equation}
learn a single monotone triangular map in this standardized space,
\begin{equation}
  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
  \label{eq:map-direction}
\end{equation}
and sample via $S^{-1}: z \mapsto u \mapsto x$. All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated. The pullback identity \eqref{eq:pullback} remains the objective we optimize.

Reported log-densities on the original scale apply the diagonal affine correction
\begin{equation}
  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
  \label{eq:affine-correction}
\end{equation}
so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable. We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.

The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}. The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}. Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.

\section{Triangular Transport Maps}

With standardized coordinates fixed, we describe the transports we fit directly.

\subsection{Separable and Cross-Term Components}

A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.

\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
  \qquad
  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
  \label{eq:methods-separable}
\end{equation}
so the log-Jacobian term depends only on $u_k$. The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.

\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
\begin{equation}
  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
  \qquad
  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
  \label{eq:methods-cross}
\end{equation}
so the log-Jacobian depends on both $u_k$ and the context. This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.

Two examples highlight the contrast. If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction. If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.

\subsection{Monotone Parameterizations}

We rely on two monotonicity mechanisms that mirror the structures above.

\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients. This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.

\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$. The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence. The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.

\subsection{Basis Choices and Tail Behaviour}

Both parameterizations combine global and local bases with explicit tail control:
\begin{itemize}
  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references. We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows. Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features. In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context. We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
\end{itemize}

\subsection{Ordering, Sparsity, and Objectives}

Triangular structure is anisotropic: the variable ordering matters. The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}. We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible. The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.

With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$. Optimizing the pullback likelihood therefore amounts to minimising the separable objective
\begin{equation}
  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
  \label{eq:ttm-objective}
\end{equation}
which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives. Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.

\subsection{Safeguards for Cross-Term Training}

The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
\begin{enumerate}
  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$. This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$). Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty. When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
\end{enumerate}

Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth. Each step is a monotone root-find with robust bracketing thanks to tail linearization. Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block. Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.

\section{Transformation Random Forests as Transport}

Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}. A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}. The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}. Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.

Under the ubiquitous additive predictor implementation,
\begin{equation}
  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
  \label{eq:methods-trtf-additive}
\end{equation}
so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$. The induced transport is separable; context shifts location but cannot change shape. Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion. Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.

The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor. This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?

\section{Copula Baselines}

Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
\begin{itemize}
  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$. The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$. This scales gracefully to high $K$ but enforces elliptical dependence.
  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian. This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
\end{itemize}
These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.

\section{Evaluation Protocol}

We evaluate all models under a common rubric.

\subsection{Log-Likelihoods and NLL}

For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$. The test log-likelihood averages this quantity over the test split, and the NLL is its negative. For triangular models we exploit the decomposition
\begin{equation}
  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
  \label{eq:nll-decomposition}
\end{equation}
with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
\begin{equation}
  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:conditional-nll}
\end{equation}
which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.

\subsection{Calibration via PIT}

Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
\begin{equation}
  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
  \label{eq:pit}
\end{equation}
should be i.i.d. $\mathrm{Unif}(0,1)$ over the test set. We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity. For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent. Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.

\subsection{Compute Metrics and Memory}

We record wall-clock training time and per-sample evaluation time on the test set. For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature. TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation. Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation. Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries. All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.

\subsection{Defaults and External Baselines}

To keep the main text readable we adopt consolidated defaults. For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation. We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere. Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.

\section{Data and Preprocessing}

We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.

\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}. The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.

\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.

\section{Metrics and Baselines}

\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).

\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
\begin{equation}
  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:eval-conditional-nll}
\end{equation}
which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.

\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}. Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion. For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.

\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g. TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.

\section{Robustness and Sensitivity}

All experiments use deterministic seeds; we average results over multiple runs and report standard errors. We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals. Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}. We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.

\section{Empirical Results}

>>>>

% Checklist: enforce US spelling, clarify Jacobian phrasing, remove empty sections.
This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling program. Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable. Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space. That alignment keeps objectives, diagnostics, and reported log-densities interoperable.

% Checklist: define model abbreviations for concise tables
\textbf{Model abbreviations.} For the four-dimensional autoregressive study we abbreviate the oracle fits as True-Marg for the marginal reference and True-Joint for the conditional reference. Transformation Random Forests appear as TRTF. This shorthand matches the axis-parallel implementation (TRTF) used in our experiments. We write the marginal triangular transport as TTM-Marg, the separable variant as TTM-Sep, and the copula mixture baseline as Copula. We use these labels across tables and figures that report this study.

\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}\label{sec:setup}

% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models. All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction. We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.

We standardize features with training-split statistics only. Equation~\eqref{eq:transport-standardise} defines $u = T_{\mathrm{std}}(x) = (x-\mu)\oslash \sigma$ with $\sigma_k > 0$. We evaluate $\log \pi_U(u)$ through the pullback identity in Equation~\eqref{eq:transport-pullback}, apply the triangular factorization from Equation~\eqref{eq:transport-det}, then convert to $\log \pi_X(x)$ using the diagonal correction in Equation~\eqref{eq:transport-affine}. We report average test negative log-likelihoods (NLL) in nats. Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the standardized pipeline shared by transport maps, Transformation Random Forests, and copulas.

We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits. Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures. The canonical four-dimensional ordering is $(1,2,3,4)$. Robustness to ordering is assessed by averaging over all $4! = 24$ permutations. We adopt the natural column order for real datasets. Transformation Random Forests are the axis-parallel implementation denoted TRTF throughout. We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability. Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.

The Half-Moon dataset provides a curved, bimodal joint in $K=2$. We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$. For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$. For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$. The observed $X = m(\Theta) + \varepsilon$. The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$. Figure~\ref{fig:halfmoon-panels} shows representative contour plots at $n=250$, which align with this generator. Table~\ref{tab:halfmoon-nll} reports the corresponding NLLs.

The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality. The first coordinate is $X_1 \sim \mathcal{N}(0,1)$. The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$. The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$. We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$. The mixing weight is $w = \sigma(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\sigma(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$. The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$. We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \sigma(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$. The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.

% Checklist: summarise generator config, define softmax usage.
\begin{table}[t]
  \centering
  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study. The softplus transform enforces positive rates, shapes, and scales.}
  \label{tab:autoregressive-config}
  \begin{tabular}{lll}
    \hline
    Coordinate & Distribution & Parameter mapping \\
    \hline
    $X_1$ & Normal & Fixed $\mathcal{N}(0,1)$ \\
    $X_2 \mid X_1$ & Exponential & $\mathrm{rate} = \mathrm{softplus}(X_1)$ \\
    $X_3 \mid X_{1:2}$ & Beta & $\mathrm{shape}_1 = \mathrm{softplus}(X_2)$; $\mathrm{shape}_2 = \mathrm{softplus}(X_1)$ \\
    $X_4 \mid X_{1:3}$ & Gamma & $\mathrm{shape} = \mathrm{softplus}(X_3)$; $\mathrm{scale} = \mathrm{softplus}(X_2)$ \\
    \hline
  \end{tabular}
\end{table}

We apply the softplus transform $\mathrm{softplus}(t) = \log(1 + \exp(t))$ to map unconstrained predictors to strictly positive distribution parameters. Mixture weights use the logistic gate $\sigma(a)$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one. For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$. This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.

Table~\ref{tab:autoregressive-config} lists the intended marginal families by dimension, and Tables~\ref{tab:autoregressive-perm} and~\ref{tab:autoregressive-perm-avg} summarize the permutation and sample-size studies used later in this chapter.

The MINIBOONE benchmark follows the published preprocessing to ensure comparability with flow-based baselines. We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes. We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features. We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables. Section~\ref{sec:realdata} records these steps and provides the dataset context. Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.

Additional UCI datasets appear only when we retain them for real-data context. POWER keeps household electricity attributes after jittering the minute-of-day encoding, dropping the calendar date and reactive-power column, and adding uniform noise to break ties. GAS keeps the \texttt{ethylene\_CO} subset, treats the series as i.i.d., removes strongly correlated attributes, and retains an eight-dimensional representation. HEPMASS keeps only the positive class from the ``1000'' split and discards five variables with repeated values to avoid density spikes. These preprocessing steps follow the same train-only standardization and reporting conventions described above. Section~\ref{sec:realdata} provides the corresponding background and positions these datasets within our evaluation.

All models use the same standardized frame and direction for evaluation, which keeps objectives, diagnostics, and reported quantities interoperable across triangular transports, TRTF, and copula baselines. This alignment is necessary for the conditional decompositions, probability integral transform (PIT) calibration checks, and compute summaries presented later in this chapter.

\section{Models and Implementation}\label{sec:models-implementation}

% Checklist: align references, enforce US spelling, ensure active voice.
This section specifies the estimators and implementation details that keep likelihoods, calibration, and compute directly comparable across models. All estimators share the transport direction $S:u \rightarrow z$, operate in standardized coordinates, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.

We implement separable lower-triangular transport maps denoted TTM-Sep. Component $k$ decomposes into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k) > 0,
  \label{eq:ttm-separable-def}
\end{equation}
so the Jacobian contribution depends only on $u_k$. The structure yields linear per-sample complexity in $K$ and exact inversion by back-substitution.

We minimize the Gaussian pullback objective induced by the shared reference,
\begin{equation}
  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
  \label{eq:ttm-separable-loss}
\end{equation}
which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}. The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions. We solve the regularized problem with bound-constrained optimization and enforce monotonicity by construction.

We construct $h_k$ with monotone one-dimensional bases that combine identity, integrated sigmoids, softplus-like edge terms, and integrated radial basis functions. Nonnegativity constraints on the derivative coefficients guarantee $h_k'(u_k) \ge 0$. We linearize tails to stabilize likelihoods as $|u_k|$ grows. Ridge penalties apply to all basis coefficients, and optional sparsity penalties shrink context shifts when multicollinearity inflates variance. During training and evaluation we clip log-derivatives to $[-H, H]$ to avoid numerical overflow in the Jacobian sum; the bound $H$ is tuned on the validation split.

We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood. This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes. Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}. When heuristics are applied, we use the identity ordering or a Cholesky-pivoted ordering with optional Gaussianization, as exposed by the shared transport utilities. Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.

We reference a cross-term variant, denoted TTM-X, only to delimit scope. The variant augments the separable component with low-rank interactions,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj}\,q_j(u_j)\,r_k(u_k),
  \label{eq:ttm-cross}
\end{equation}
where $q_j$ and $r_k$ are monotone features and constraints ensure $\partial_{u_k} S_k(u_{1:k}) > 0$. We exclude TTM-X from headline tables because the interactions alter identifiability and complicate calibration. The definition clarifies the naming used in the synthetic analyses.

We implement Transformation Random Forests with additive predictors and denote the model TRTF. Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest. The induced triangular component is
\begin{equation}
  S_k(u_{1:k}) = \Phi^{-1}\!\big(\widehat{F}_k(u_k \mid u_{1:k-1})\big),
  \label{eq:trtf-transport}
\end{equation}
and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$. Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame. Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.

We keep copulas as dependence baselines with explicit scope. We fit only low-dimensional nonparametric copulas for $K \le 3$, using probit-transformed pseudo-observations and kernel density estimation on the Gaussian scale before mapping back to the unit cube with the appropriate Jacobian. The independence baseline evaluates the product of fitted marginals. We omit a Gaussian copula from the main experiments to preserve consistency with the low-$K$ nonparametric dependence analyzed in Section~\ref{sec:realdata}.

We adopt a single inversion and evaluation convention across estimators. Training, Jacobians, and conditional evaluations occur in standardized coordinates. Sampling draws $z \sim \mathcal{N}(0, I)$, applies $S^{-1}$ by back-substitution, and converts to the original scale with the stored affine parameters. This convention prevents mixed objectives and keeps all reported quantities interoperable.

We ensure reproducibility and comparability with fixed seeds, cached standardization parameters, and shared reporting utilities. Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction, nonparametric copulas, marginal and separable triangular maps, and the shared transport core that implements ordering, bases, derivatives, and evaluation. The appendix also records optimizer choices, timing hooks, and object layouts used in the experiments.

\section{Evaluation Metrics and Protocol}\label{sec:evaluation-protocol}

% Checklist: align references, keep active voice, reuse existing notation.
This section defines the metrics and procedures applied across all models so likelihoods, calibration, and compute remain directly comparable. We evaluate every estimator in standardized coordinates, apply the triangular determinant, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.

We distinguish pointwise log density from dataset averages. The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative. Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity, and all values appear in nats with consistent precision. Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}. Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.

Triangular models exploit the separable pullback in standardized coordinates. With $u = T_{\mathrm{std}}(x)$, the log density decomposes as
\begin{equation}
  \log \hat{\pi}_U(u) = \sum_{k=1}^{K} \Big[ \log \varphi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
  \label{eq:evaluation-triangular}
\end{equation}
so the determinant factorization in Equation~\eqref{eq:transport-det} yields linear per-sample cost in $K$. In plain language, the model checks how Gaussian each transformed coordinate looks, then adds the exact log-Jacobian contribution from its one-dimensional derivative. The affine correction in Equation~\eqref{eq:transport-affine} converts $\log \hat{\pi}_U$ to $\log \hat{\pi}_X$ for reporting.

We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
\begin{equation}
  \mathrm{NLL}_k = -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:evaluation-conditional-nll}
\end{equation}
and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction. Copulas lack a unique triangular factorization, so we report only their joint NLL. Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains. These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame.
